---
title: "Decision tree from SPSS output"
author: "Sarah Gibbs"
date: "July 15, 2025"
output: html_document
editor_options: 
  chunk_output_type: console
---
# This document is intended for improving the look of a classification tree that has already been rendered in SPSS. You can use the default display settings that SPSS provides and simply export it as a PMML file (double click on the tree to open the editor, then click File > Export PMML).
# While the function is handy and easy to use, depending on your desire (and the size and depth of your tree) you may have to do some additional image processing after rendering. I recommend Inkscape (open-source). For example, due to package limitations (as well as the limitations of my own coding abilities) certain dynamic labels cannot be rendered in the same function as the static lines that connect branches (bottom half of code).
# Load packages
```{r}
library(ggplot2)
library(XML)
library(patchwork)
library(data.tree)
library(RColorBrewer)
library(grid)
library(cowplot)
library(stringr)
```

# Load data
```{r}
pmml <- xmlParse("C:\\Users\\sgibb\\OneDrive\\Documents\\Scyphers Lab\\Reef Fish Extension\\Survey Analysis\\DecisionTree.xml") # change this to your own file directory
ns <- c(pmml = "http://www.dmg.org/PMML-4_3")
```

# Get variable names and labels
```{r}
# Extract labels from DataDictionary (Variable names and class labels)
extract_labels <- function(pmml, ns) {
  data_fields <- getNodeSet(pmml, "//pmml:DataField", namespaces = ns)
  
  # Get predicted field name
  predicted_field_node <- getNodeSet(pmml, "//pmml:MiningField[@usageType='predicted']", namespaces = ns)[[1]]
  predicted_field_name <- xmlGetAttr(predicted_field_node, "name")

  # Create result lists
  class_labels <- list()
  variable_labels <- list()
  value_labels <- list()

  # Iterate through each variable in dictionary
  for (field in data_fields) {
    field_name <- xmlGetAttr(field, "name")
    display_name <- xmlGetAttr(field, "displayName", default = field_name)

    # Store variable name
    variable_labels[[field_name]] <- display_name

    # Store value labels (if they exist)
    value_nodes <- getNodeSet(field, "./pmml:Value", namespaces = ns)
    if (length(value_nodes) > 0) {
      value_map <- sapply(value_nodes, function(x) {
        val <- xmlGetAttr(x, "value")
        disp <- xmlGetAttr(x, "displayValue", default = val)
        disp
      })
      names(value_map) <- sapply(value_nodes, function(x) xmlGetAttr(x, "value"))
      value_labels[[field_name]] <- value_map
      
      # Save class labels if variable is predicted
      if (field_name == predicted_field_name) {
        class_labels <- value_map
      }
    }
  }

  return(list(
    class_labels = class_labels,
    variable_labels = variable_labels,
    value_labels = value_labels
  ))
}

labels <- extract_labels(pmml, ns)
class_labels <- labels$class_labels
variable_labels <- labels$variable_labels
value_labels <- labels$value_labels
```

# Parse the decision tree from PMML file
```{r}
# Get predicted variable name
get_title <- function(pmml, ns) {
  # 1. Find the predicted field name
  predicted_field_node <- getNodeSet(pmml, "//pmml:MiningField[@usageType='predicted']", namespaces = ns)[[1]]
  predicted_field_name <- xmlGetAttr(predicted_field_node, "name")

  # 2. Look it up in the DataDictionary
  xpath <- paste0("//pmml:DataField[@name='", predicted_field_name, "']")
  data_field_node <- getNodeSet(pmml, xpath, namespaces = ns)[[1]]
  display_name <- xmlGetAttr(data_field_node, "displayName")

  return(display_name)
}

# Categorical variable splits are nested within the model, this makes sure they are being referenced
get_first_split <- function(xml_node, ns) {
  preds <- getNodeSet(xml_node, ".//pmml:SimplePredicate", namespaces = ns)
  if (length(preds) > 0) {
    pred <- preds[[1]]
    field <- xmlGetAttr(pred, "field")
    op <- xmlGetAttr(pred, "operator")
    val <- xmlGetAttr(pred, "value")
    return(paste(field, op, val))
  } else {
    return(NA)
  }
}

parse_node <- function(xml_node, ns, variable_labels, value_labels, class_labels) {
  node_id <- xmlGetAttr(xml_node, "id")
  score <- as.character(xmlGetAttr(xml_node, "score"))
  count <- as.numeric(xmlGetAttr(xml_node, "recordCount"))
  
  split_raw <- get_first_split(xml_node, ns)
  if (!is.na(split_raw)) {
    parts <- strsplit(split_raw, " ")[[1]]
    var <- parts[1]
    op <- parts[2]
    val <- parts[3]

    var_display <- variable_labels[[var]]
    val_display <- value_labels[[var]][[val]]
    if (is.null(val_display))
      val_display <- val # If variable is continuous, raw numeric string will be added

    op_symbol <- switch(op,
                      "lessOrEqual" = "≤",
                      "greaterThan" = ">",
                      "equal" = "=",
                      op) 

    split <- list(
      variable = var_display, # Question text
      value = paste(op_symbol, val_display) # Split response
    )
  } else {
    split <- NULL
  }

  # Get score distribution
  dist_nodes <- getNodeSet(xml_node, "./pmml:ScoreDistribution", namespaces = ns)
  classes <- sapply(dist_nodes, function(x) xmlGetAttr(x, "value"))
  freqs <- sapply(dist_nodes, function(x) as.numeric(xmlGetAttr(x, "recordCount")))
  probs <- round((freqs / sum(freqs)) * 100, 1)

  # Parse children
  children_nodes <- getNodeSet(xml_node, "./pmml:Node", namespaces = ns)
  children <- lapply(children_nodes, function(x) parse_node(x, ns, variable_labels, value_labels, class_labels))

  list(
    node_id = node_id,
    parent_id = NULL,
    score = score,
    count = count,
    label = paste0(
      #class_labels[score], # If you want to add predicted categories for each donut
      "\nN=", count),
    split = split,
    probs = data.frame(Class = classes, Freq = probs),
    children = children
  )
}

# Get root node
tree_root <- getNodeSet(pmml, "//pmml:TreeModel/pmml:Node", namespaces = ns)[[1]]
tree_data <- parse_node(tree_root, ns, variable_labels, value_labels, class_labels)

# Get title 
tree_title <- get_title(pmml, ns)
```

# Donut plot for each node
```{r}
donut_plot <- function(probs_df, class_labels, palette, count, show_legend = F) {

  probs_df$ClassLabel <- factor(class_labels[as.character(probs_df$Class)], levels = unname(class_labels))
  
  p <- ggplot(probs_df, aes(x = 2, y = Freq, fill = ClassLabel)) +
    geom_bar(stat = "identity", color = "white", width = 1) +
    coord_polar("y", start = 0) +
    geom_text(aes(label = paste0(Freq, "%")),
              position = position_stack(vjust = 0.5)) +
    xlim(0.5, 2.5) +
    theme_void() +
    annotate(geom = 'text', x = 0.5, y = 0, label = paste0("N = ", count)) + # add sample size to center
    scale_fill_manual(name = "Satisfaction Level", values = palette, drop = F, 
                      guide = guide_legend(reverse = T)) +
    theme(legend.position = if (show_legend) "right" else "none",
      plot.title = element_text(size = 10)
    )
  
  cowplot::ggdraw(p)
}
```

# Plot all nodes in tree layout
```{r}
render_tree <- function(node, class_labels, palette, title = NULL, depth = 0) {
  
  is_root <- depth == 0

  donut <- donut_plot(node$probs, class_labels, palette, node$count, show_legend = is_root)

  # If no children, return donut (no split at leaves)
  if (length(node$children) == 0) {
    return(donut)
  }

  # 1. Render all child donuts with their *split value labels* ABOVE
  child_with_value_labels <- lapply(node$children, function(child) {
    # Get label like "< Very Satisfied"
    value_label <- if (!is.null(child$split$value)) child$split$value else ""
    value_label_plot <- cowplot::ggdraw() + cowplot::draw_label(
      value_label, fontface = "italic", size = 9, hjust = 0.5
    )
    donut_plot <- render_tree(child, class_labels, palette, title = NULL, depth = depth + 1)
    cowplot::plot_grid(value_label_plot, donut_plot, ncol = 1, rel_heights = c(0.12, 1))
  })

  children_row <- cowplot::plot_grid(plotlist = child_with_value_labels, nrow = 1, align = "h")

  # 2. Add split variable label ABOVE the entire row of children (but only if not root)
  if (!is_root && !is.null(node$children[[1]]$split$variable)) {
    split_variable_label <- cowplot::ggdraw() + cowplot::draw_label(
      str_wrap(node$children[[1]]$split$variable, width = 45),
      fontface = "bold",
      size = 11,
      hjust = 0.5
    )
    children_row <- cowplot::plot_grid(split_variable_label, children_row, ncol = 1, rel_heights = c(0.15, 1))
  }

  # 3. Add title above root node only
  if (is_root) {
  # Title above everything
  title_label <- cowplot::ggdraw() + cowplot::draw_label(title, fontface = "bold", size = 13, hjust = 0.5)

  # Root split variable label (from first child)
  if (!is.null(node$children[[1]]$split$variable)) {
    root_split_label <- cowplot::ggdraw() + cowplot::draw_label(
      str_wrap(node$children[[1]]$split$variable, width = 45),
      fontface = "bold", size = 11, hjust = 0.5
    )
    final_stack <- cowplot::plot_grid(
      title_label, donut, root_split_label, children_row,
      ncol = 1, rel_heights = c(0.12, 1, 0.15, 1)
    )
  } else {
    final_stack <- cowplot::plot_grid(title_label, donut, children_row, ncol = 1, rel_heights = c(0.12, 1, 1))
  }

} else {
  # Regular subtree layout
  final_stack <- cowplot::plot_grid(donut, children_row, ncol = 1, rel_heights = c(1, 1))
}

  return(final_stack)
}
```

# Display tree
```{r}
# Specify palette
palette <- setNames(brewer.pal(5, "RdYlBu"), unname(class_labels)) # links colors to labels, so if a class is missing color order is still maintained

tree_plot <- render_tree(tree_data, class_labels, palette, tree_title)
tree_plot
```

# Export tree
```{r}
# If you are doing any post processing with an image software (e.g., Inkscape), .svg files should be used
ggsave("tree_plot.svg", width = 18, height = 12, units = "in")
```

# Below, the tree will render with connectors for each branching node, but you have less flexibility with generating labels.

# Build layout with coordinates
```{r}
flatten_tree_with_coords <- function(node, x = 0, y = 0, parent_id = NULL, level = 0, spacing = 1) {
  node_info <- list(
    id = node$node_id,
    parent = parent_id,
    x = x,
    y = -level,
    node = node
  )

  if (length(node$children) == 0) {
    return(list(node_info))
  }

  # Calculate child positions
  n <- length(node$children)
  width <- spacing * (n - 1)
  child_positions <- seq(x - width / 2, x + width / 2, length.out = n)

  children <- unlist(
    mapply(function(child, cx) {
    flatten_tree_with_coords(child, x = cx, y = y - 1, parent_id = node$node_id, level = level + 1, spacing = spacing / 1.5)
  }, node$children, child_positions, SIMPLIFY = FALSE),
  recursive = FALSE)

  return(c(list(node_info), children))
}

```

# Plot all nodes and connectors
```{r}
render_tree_with_connectors <- function(flat_tree, class_labels, palette, title = NULL) {

# Make a named list for easy lookup by node ID
node_lookup <- setNames(flat_tree, sapply(flat_tree, function(e) e$id))

  # Build node plots
  plots <- lapply(flat_tree, function(entry) {
    node <- entry$node
    depth <- abs(entry$y)
    is_root <- is.null(entry$parent)
    show_legend <- is_root
    label <- paste0("N = ", node$count)

    p <- donut_plot(node$probs, label, class_labels, palette, show_legend = show_legend)

    # Add split value above, if present and not root
    if (!is_root && !is.null(node$split$value)) {
      label_layer <- ggplot() +
        geom_label(aes(x = 0.5, y = 0.5, label = node$split$value),
                   fill = "white", color = "black", size = 3, label.size = NA,
                   fontface = "italic") +
        theme_void() +
        xlim(0, 1) + ylim(0, 1)
        #cowplot::ggdraw() + cowplot::draw_label(
        #node$split$value,
        #fontface = "italic", size = 9, hjust = 0.5,
        #fill = "white", label.size = NA  # white background
      #)
      p <- cowplot::plot_grid(label_layer, p, ncol = 1, rel_heights = c(0.15, 1))
    }

    return(p)
  })

  # Compute layout extents
  min_x <- min(sapply(flat_tree, function(n) n$x))
  max_x <- max(sapply(flat_tree, function(n) n$x))
  min_y <- min(sapply(flat_tree, function(n) n$y))
  max_y <- max(sapply(flat_tree, function(n) n$y))

  # Create empty base canvas
  base <- ggplot() + xlim(min_x - 1, max_x + 1) + ylim(min_y - 1, max_y + 1) + theme_void()

  # Add connector lines
  connectors <- do.call(rbind, lapply(flat_tree, function(entry) {
  if (!is.null(entry$parent)) {
    parent <- node_lookup[[entry$parent]]
    if (!is.null(parent)) {
      data.frame(
        x = parent$x,
        y = parent$y - 0.1,
        xend = entry$x,
        yend = entry$y + 0.1
      )
    }
  }
}))
  
  base <- base + geom_segment(
  data = connectors,
  aes(x = x, y = y, xend = xend, yend = yend),
  linewidth = 0.5,
  color = "gray30",
  inherit.aes = FALSE
)

  # Add all donut plots as annotation_custom grobs
  for (i in seq_along(flat_tree)) {
    entry <- flat_tree[[i]]
    p <- plots[[i]]
    g <- ggplotGrob(p)
    base <- base + annotation_custom(g, xmin = entry$x - 0.5, xmax = entry$x + 0.5,
                                     ymin = entry$y - 0.5, ymax = entry$y + 0.5)
  }

  # Add title (once)
  if (!is.null(title)) {
    title_row <- cowplot::ggdraw() + cowplot::draw_label(title, fontface = "bold", size = 14, hjust = 0.5)
    return(cowplot::plot_grid(title_row, base, ncol = 1, rel_heights = c(0.12, 1)))
  }

  return(base)
}
```

# Display tree
```{r}
# Specify palette
palette <- setNames(brewer.pal(5, "RdYlBu"), unname(class_labels)) # links colors to labels, so if a class is missing color order is still maintained

flat <- flatten_tree_with_coords(tree_data)
tree_plot <- render_tree_with_connectors(flat, class_labels, palette, title = tree_title)
tree_plot
```

# Export tree
```{r}
# If you are doing any post processing with an image software (e.g., Inkscape), svg files should be used
ggsave("tree_plot.svg", width = 18, height = 12, units = "in")
```